Characterizing Stellar Streams by ML

### 2023-08-26 src/tex/ms.tex
[showyourwork] overleaf sync

\section{Introduction} \label{sec:intro}
    Stellar streams are the disrupted remnants of globular clusters and satellite galaxies, and their census has grown extensively in the Milky Way thanks to all-sky astrometric missions like {\it Gaia} and low-surface brightness photometric surveys (\JN{cite}). The formation of streams can be traced back to a progenitor cluster (i.e., a satellite galaxy or globular cluster), whose stars are gradually lost to the tidal field of the host galaxy and extend along a series of similar orbits producing tidal tails, which can sometimes span many tens of degrees across the sky. This orbital proximity of stars belonging to a stream makes them sensitive tracers of the mass distribution of galaxies, sourced by both the baryonic and dark matter components.

    The average properties of a stream (e.g., the mean position and velocity track of the tidal tails) can be used to reconstruct the global dark matter distribution of the stellar halo (e.g., \JN{cite}), while local variations in the stream track and surface density encode information about the stream's encounters with baryonic structures such as (e.g.) the stellar bar (\JN{cite}), and non-luminous structures like dark matter subhalos (\JN{cite}). Importantly, these galactic components span orders of magnitude in mass, highlighting the importance of characterizing a population of stellar streams both in morphology and kinematics.

    While several methods have been developed to model stream properties as a function of global halo properties (e.g., flattening and orientation;\JN{cite}) and local substructures (e.g., the bar's pattern speed and the mass of subhalo perturbers; \JN{cite}), these methods and insights often rely on the existence of a homogenized stream-tracer population that characterizes the stream across each of the observable phase-space dimensions. Additionally, because even the fine-grain properties of streams are influenced by the detailed mass distribution of galaxies, it is important to propagate errors both in measurement uncertainty and membership uncertainty (i.e., which stars belong to the stream versus the background) when attempting to model the galaxy. Otherwise, a stream that has been poorly characterized could bias constraints on (e.g.) the total mass of the galaxy, the dark matter halo shape, and the mass of subhalo perturbers. In order to generate more precise constraints on the structure of the galaxy, it is therefore imperative that a statistically sound and homogenized catalog of stellar streams is produced so that theoretical models for the gravitational potential of the galaxy across a large range of spatial scales can be constrained.

    While streams are well described in (e.g.) action angle coordinates or $N-$body simulations, these methods relying on prior assumptions about the galaxy and its underlying potential. Therefore, it is important to devise data-driven methods to characterize stellar streams without appealing to strong prior assumptions about the galaxy's structure.

    There are a few previous works which have modeled stellar streams in this context. \citet{2022MNRAS.514.1757P} developed a data-driven spline-based model for modeling the photometric qualities of previously identified streams. Their method works by fitting the stream's stellar surface density with a series of splines, capturing both surface brightness variations along the stream and changes in the position of the stream track and width. Their method also utilize a simultaneous ischorone fit to the stream, from which a distance modulus can be derived by assuming that the entire stream originates from the same stellar population. While this method is very successful at characterizing the photometric properties of a stream, it does not consider the kinematic dimensions which have become increasing well-measured with successive data releases from {\it Gaia}, as well as targeted observations. From a population of RR Lyrae stars, \citet{2019AJ....158..223P} utilized a mixture model to fit the globular cluster stream Pal-5 in the space of proper motions and heliocentric distances. The mixture modeling approach enables a simultaneous fit to both the stream and background density, enabling better separation of a coherent stream from an otherwise noisy background distribution. However, the track of the stream in on-sky coordinates is not fit. The method identified 27 RR Lyrae stars consistent with being members of the stream. While useful for determining the average properties of the stream (e.g., its mean distance track), a limited population of RR Lyrae stars are not sufficient to capture the fine-grained properties of streams expected to inform constraints on (e.g.) a population of perturbers in the galaxy. \JN{are there other methods?}

    Other methods for extracting stellar streams from the data include matched-filter approaches, where over-densities in color-magnitude space are inspected as possible clusters undergoing tidal stripping \JN{cite the many key papers using this method}. While the matched-filter technique has been extremely successful at stream discovery, it is not equipped to characterize streams at the level of detailed necessary to inform precise constraints on the potential of the galaxy and its substructure.

    In order to address the need for a flexible model of stellar streams both in photometric and kinematic spaces, in this work we develop a new method for characterizing streams and quantifying the membership probability of possible stream stars. Our approach provides several key advantage over previous stream modeling efforts, specifically in its ability to jointly model the kinematic and photometric properties of streams simultaneously. Furthermore, our method is well-suited to fitting streams with only partial or incomplete phase-space or photometric observations, enabling a full characterization of streams insofar as the available data allows.

@@

        % Observational errors complicate this picture. Instead of a certain membership calculation, we must build a probabilistic model, assigning a likelihood that an observed star falls along the stream's orbit.

@@

        The paper is organized as follows.
        In \autoref{sec:method} we introduce our method for characterizing stream tracks and computing stellar membership.
        We discuss the data in \autoref{sub:framing_the_data} and build a probabilistic framework to model the stream and background in \autoref{sub:likelihood_setup}-\autoref{sub:priors}.
        In \autoref{sec:results_mock} we apply our method to simulated data to demonstrate the model with a known ground truth dataset,
        then to the stellar streams \stream{GD-1} (\autoref{sec:results_gd1}) and \stream{Pal\,5} (\autoref{sec:results_pal5}).
        We compare our approach to characterizing stellar streams to other methods in \autoref{sec:comparison}.
        We conclude in \autoref{sec:conclusions}, discussing the results of this work and future directions.

@@

    We now introduce our method for characterizing stellar stream tracks and computing stream membership likelihoods from N-D phase-space measurements.
    % TODO: rephrase following sentence
    We develop a flexible Bayesian model framework, that allows arbitrary numbers of phase-space dimensions to be incorporated in the model and permits missing phase-space observations. 
    Below, we offer a concise overview of our approach, followed by a more in-depth technical explanation in subsequent sections.
    We start, in \autoref{sub:framing_the_data}, by transforming the astrometric data into
    a stream-oriented on-sky reference frame ($\phi_1, \phi_2, \parallax, \mu_{\phi_1}^*, \mu_{\phi_2}, v_r)$, for each stream. In this frame, 
    the stream track lies along $\phi_1$, and the stream density may be described by distributions in the other parameters as injective functions of $\phi_1$.
    % TODO: ⬆ I think a few more sentences explaining why this is important, feature variations, etc?
    % However, previously we discussed how stream will vary along the track, with changes in the width and possible gaps.
    % Also observational effects, e.g the distance gradient changes the on-sky width (should this be mentioned earlier?)
    Likewise, the background of the Galactic field may be described with $\phi_1$-conditioned distributions.
    % TODO: ⬆ I think this should be expanded.
    Non-astrometric coordinates, e.g. photometry and metallicity, are also modelled along the stream and background as functions of $\phi_1$.

@@

    \subsection{Framing the Data} \label{sub:framing_the_data}

@@

    \subsection{Likelihood Setup}\label{sub:likelihood_setup}

@@

        The mixture coefficients $\qth{f}$ are normalized s.t.
        \begin{equation}
            \sum_{\mrm{q}} \qth{f}(\phi_1) = 1 \qquad \forall \phi_1
        \end{equation}
        In practice this is enforced by defining
        \begin{equation} \label{eq:mixture_weight_normalization}
            f_{b}(\phi_1) = 1 - \sum_{\rm{q}\in \mbs{I_Q}\backslash \{b\}} \qth{f}(\phi_1),
        \end{equation}
        the summation over all models except for some ``background" index $b \in \mbs{I_Q}$.
        We discuss the mixture weights further in \autoref{sub:mixture_weight}.

@@

    \subsection{Missing Phase-Space Observations} \label{sub:missing_data}

@@

    \vspace{10pt}
    \subsection{Determination of the Mixture Weight} \label{sub:mixture_weight}
        The mixture weight $f_q(\phi_1)$ as used in \autoref{eq:general_model} is characterized as the fraction of stars belonging to the mixture component $q$ over an increment $[\phi_1, \phi_1 + \dif{\phi_1}]$. Thus, the mixture weight encodes information about the linear density of the stream, which has been shown to depend on substructure in the galaxy (CITE-x). The dependence of the mixture weight on the linear density is captured by the expression
        \begin{equation}\label{eq:fraction_param}
            f_q\left(w_{i,0}\right) = \frac{N_q\left(\phi_1\right)}{N_{\rm tot}\left(\phi_1\right)},
        \end{equation}
        where $N_q\left(\phi_1\right)$ is the number of stars in component $q$ within an interval $[\phi_1, \phi_1 + \dif{\phi_1}]$, and 
        $N_{\rm tot}$ is the total number of stars in the same band (including those in component $q$). The linear  density for component $q$ is then entirely specified by the function $N_q(\phi_1)$.

@@

    \subsection{Astrometric Model} \label{sub:astrometric_model}

        \subsubsection{On-Stream} \label{ssub:astrometric_model_on_stream}

@@

            As a MDN, we implement the Gaussian distribution parameters with a multi-layered feed-forward neural network. The number of layers required depends on the maximum dimensionality of the Gaussian, and whether the neural network employs regularization techniques like dropout. We use alternating layers of linear and tanh units. Details of the network architecture are discussed in \autoref{sub:model_implementation}.

@@

        \subsubsection{Off-stream} \label{ssub:astrometric_model_off_stream}

            We now outline the off-stream astrometric model. We take a similar approach to the on-stream astrometric model discussed in \autoref{ssub:astrometric_model_on_stream}, though we typically do not characterize the backgrounds with a Gaussian distribution. Instead, we utilize a range of distributions (i.e., exponential, skew-normal) discussed in \autoref{app:distributions}. The parameters of the user-specified background distributions are themselves neural network outputs, such that any $\phi_1$ ``slice" of the background should be characterized by some analytic density, but the evolution of the backgrounds with $\phi_1$ can be complex. Details of the network architecture are discussed in \autoref{sub:model_implementation}.

@@

    \subsection{Photometric Model} \label{sub:photometric_model}

        \subsubsection{On-Stream} \label{sub:photometric_model_on_stream}
            In photometric coordinates we model the stream as a single-population isochrone with the possibility of a non-zero distance gradient. We accomplish this by modeling the distance of the isochrone as a neural network output, parametrized as a function of $\phi_1$. The result is a distance track along the stream, estimated from the variation in its isochrone with $\phi_1$.

@@

            The second term in \autoref{eq:isochrone_prior_components} -- $\prior_{obs}(\nth{\mbs{m}} | \phi_1)$ -- encodes all observational effects in the photometric dataset. Some effects are easily modeled, e.g. instruments have well-characterized magnitude limits, both faint and bright. Other effects are not easily modeled, e.g, complex location-dependent completeness issues \citep{GaiaCompleteness}.
            Accounting for the completeness is very challenging. In practice we avoid
            the contribution of completeness to $\prior_{obs}$, masking data for which it is relevant. \Gaia, for instance, is complete from $12 < G < 17$ and has high completeness up to $G ~20$, thus we mask $G > 20$.
            The masks are easily modeled as magnitude limits and replace the complex completeness model in $\prior_{obs}$.
            \autoref{sub:missing_data} explains how masking is implemented in the models, though in this case every feature is masked, not only a few feature dimensions. With every featured masked, the photometric model in \autoref{eq:photometric_probability} reduces from an $F_m$-dimensional Gaussian to a 0-D Gaussian, which is a null distribution and does not enter into the likelihood.
            In photometric coordinates completeness depends on a stars apparent magnitude; vied in in astrometric coordinates, the completeness cannot be described so simply. \Gaia, for instance has scanning law residuals in the density field. When photometric coordinates are masked due to completeness, we do not mask the corresponding  astrometric features, so that the stars  membership probability is determined solely by the astrometric model. This practice allows us to include stars with good astrometric measurements, not penalizing those stars for which the photometric model is insufficient.
            We examine the results to confirm that the scan pattern does not impact the
            membership distribution.

@@

            % TODO? expand this section.
            The last approach to the issue of the PDMF, and the one we take in this work, is to render it unimportant. 
            With fixed age, metallicity, and other isochrone model parameters, the primary purpose of the isochrone is to allow the model to determine the distance modulus of the stream. For this, only a portion of the stream need be modeled.
            For streams with observable main sequence turn-offs, like \stream{GD-1} in \gaia, this portion of the isochrone has both many stars and a relatively small range in masses. Thus, we restrict the mass range of the isochrone such that we are modeling only a region of interest over which the PDMF does not contribute to the final result.
            % TODO? work in something to the effect of:
            % The PDMF doesn't have an enormous impact, regardless as the isochrone model is very thin and stars close to the isochrone photometrically are generally quite far astrometrically. This is an advantange of simultaneous modeling, where errors in one model are balanced by other parts of the model.

@@

        \subsubsection{Linking to Astrometrics: the Distance Modulus and the Parallax} \label{ssub:linking_to_astrometrics}

@@

        \subsubsection{Off-stream} \label{sub:photometric_model_off_stream}
            In this section we introduce our model for the distribution of backgrounds in photometric coordinates. 
            The distribution of stars in color (i.e., $g-r$) and magnitude (i.e., $g$) is a complicated function of location in the galaxy \TODO{(e.g. [cite])}. Even within a field centered on the stream of interest, the color-magnitude diagram (CMD) is a combination of stellar populations distributed over a range in distances. Because of this, we do not expect an analytic density to provide a realistic description of the data. In this work we take a data-driven approach, and represent the CMD using a normalizing flow model.

@@

            % The specific architecture of the photometric normalizing flow are discussed in \autoref{sub:model_implementation}. But
            It is easy to evaluate the log-likelihood a normalizing flow at a point $\mbs{m}$ when all feature dimensions are present. When one (or more) feature dimension is missing, however, evaluating the marginal log-likelihood involves integrals over the flow. While the procedure is straightforward it is computationally expensive. As an alternative, we conservatively set the log-likelihood to $0$, the maximum value, overemphasizing the flow's contribution to the background model. Conceptually, up-weighting the background model decreases the false-positive rate in stream membership identification, at the
            cost of increasing the false-negative rate. However, the intrinsic width of stellar streams tends to be quite small, with correspondingly small $\sigma_\mu$ (see \autoref{eq:photometric_probability}); thus the effective region of the isochrone model is well determined and we do not find the marginal log-likelihood approach to be impactful. Another approach, which we leave to a future work, is to construct normalizing flows of each marginal distribution, e.g. $q_{\rm flow}\left(\{m_i\}_{i \in \mbs{I}_F, i \neq {j, k}} | m_j, m_k, \phi_1\right)$, and select for each $\nth{\mbs{m}}$ the correct flow given the present features.

@@

    \subsection{Priors} \label{sub:priors}

        In \autoref{sub:likelihood_setup} we introduced the likelihood setup. In \autoref{sub:photometric_model_on_stream} we added priors in the form of the isochrone's PDMF and photometric observational constraints. In this section we
        introduce and develop more priors that are important for training the BMDN's parameters' neural network(s).

        \subsubsection{Parameter Bounds} \label{ssub:parameter_bounds}

@@

        \subsubsection{Region Prior} \label{sub:track_region_prior}

            In \autoref{ssub:astrometric_model_on_stream} the stream model's parameters were unconstrained. Bounds on the parameters and the neural network architecture were introduced in \autoref{ssub:parameter_bounds} to ensure the initial state of the model and network were compatible with the optimizer. Now we provide known information about physical systems, particular the track of the stream.
            Our density modeling is aimed at characterization of \textit{known} streams, not stream discovery. From previous studies, the broad track of a
            stream is known for significant portions of its length \citep[e.g. see the atlas in][]{Mateu2022}. It makes sense, then, to provide this information to the model. Care must be taken that this prior information does not dominate the model.
            % A good balance between these two goals is to have a prior that only influences the track on large scales.
            We introduce a region prior to encourage the stream model to pass through a user-defined control region, while also remaining compatible with the data.
            We refer to the prior as a ``split"-Normal prior: a Gaussian split and separated at the peak and pieced back together with a flat region connecting the halves. The un-normalized PDF is given by:
            \begin{small}
            \begin{equation}
                \pdf(\theta,\mu,\sigma,w) \propto \begin{cases} 
                   \Exp{-\frac{1}{2}\left(\frac{x-(\mu-w)}{\sigma}\right)^2} & \phantom{\mu - w <}\ x < \mu - w \\
                    1 & \mu - w \leq x < \mu + w \\
                    \Exp{-\frac{1}{2}\left(\frac{x-(\mu+w)}{\sigma}\right)^2} & \mu + w < x
                \end{cases}.
            \end{equation}\end{small}
            The piecewise PDF is smooth up to the first derivative and can be used with gradient-based optimizers.

@@

    \vspace{5pt}
    \subsection{Model Implementation, aka the code} \label{sub:model_implementation}

@@

        The models permit user-defined priors to be passed at model instantiation.
        Priors specific to a single model are contained on that model, while priors connecting two (or more) models are held on the mixture model object. As an example of the latter, see \autoref{ssub:linking_to_astrometrics}, which connects the distance component of the astrometric and photometric stream models.

>>>>

\section{Introduction} \label{sec:intro}
    %
    Stellar streams are the disrupted
    remnants of globular clusters and satellite galaxies, and their census has
    grown extensively in the Milky Way thanks to all-sky astrometric missions
    like {\it Gaia} and low-surface brightness photometric surveys (\JN{cite}).
    The formation of streams can be traced back to a progenitor cluster (i.e., a
    satellite galaxy or globular cluster), whose stars are gradually lost to the
    tidal field of the host galaxy and extend along a series of similar orbits
    producing tidal tails, which can sometimes span many tens of degrees across
    the sky. This orbital proximity of stars belonging to a stream makes them
    sensitive tracers of the mass distribution of galaxies, sourced by both the
    baryonic and dark matter components.
    % 
    The average properties of a stream (e.g., the mean position and velocity
    track of the tidal tails) can be used to reconstruct the global dark matter
    distribution of the stellar halo (e.g., \JN{cite}), while local variations
    in the stream track and surface density encode information about the
    stream's encounters with baryonic structures such as (e.g.) the stellar bar
    (\JN{cite}), and non-luminous structures like dark matter subhalos
    (\JN{cite}). Importantly, these galactic components span orders of magnitude
    in mass, highlighting the importance of characterizing a population of
    stellar streams both in morphology and kinematics. 
    % 
    While several methods have been developed to model stream properties as a
    function of global halo properties (e.g., flattening and
    orientation;\JN{cite}) and local substructures (e.g., the bar's pattern
    speed and the mass of subhalo perturbers; \JN{cite}), these methods and
    insights often rely on the existence of a homogenized stream-tracer
    population that characterizes the stream across each of the observable
    phase-space dimensions. Additionally, because even the fine-grain properties
    of streams are influenced by the detailed mass distribution of galaxies, it
    is important to propagate errors both in measurement uncertainty and
    membership uncertainty (i.e., which stars belong to the stream versus the
    background) when attempting to model the galaxy. Otherwise, a stream that
    has been poorly characterized could bias constraints on (e.g.) the total
    mass of the galaxy, the dark matter halo shape, and the mass of subhalo
    perturbers. In order to generate more precise constraints on the structure
    of the galaxy, it is therefore imperative that a statistically sound and
    homogenized catalog of stellar streams is produced so that theoretical
    models for the gravitational potential of the galaxy across a large range of
    spatial scales can be constrained. 
    %
    While streams are well described in (e.g.) action angle coordinates or
    $N-$body simulations, these methods relying on prior assumptions about the
    galaxy and its underlying potential. Therefore, it is important to devise
    data-driven methods to characterize stellar streams without appealing to
    strong prior assumptions about the galaxy's structure. 
    % 
    There are a few previous works which have modeled stellar streams in this
    context. \citet{2022MNRAS.514.1757P} developed a data-driven spline-based
    model for modeling the photometric qualities of previously identified
    streams. Their method works by fitting the stream's stellar surface density
    with a series of splines, capturing both surface brightness variations along
    the stream and changes in the position of the stream track and width. Their
    method also utilize a simultaneous ischorone fit to the stream, from which a
    distance modulus can be derived by assuming that the entire stream
    originates from the same stellar population. While this method is very
    successful at characterizing the photometric properties of a stream, it does
    not consider the kinematic dimensions which have become increasing
    well-measured with successive data releases from {\it Gaia}, as well as
    targeted observations. From a population of RR Lyrae stars,
    \citet{2019AJ....158..223P} utilized a mixture model to fit the globular
    cluster stream Pal-5 in the space of proper motions and heliocentric
    distances. The mixture modeling approach enables a simultaneous fit to both
    the stream and background density, enabling better separation of a coherent
    stream from an otherwise noisy background distribution. However, the track
    of the stream in on-sky coordinates is not fit. The method identified 27 RR
    Lyrae stars consistent with being members of the stream. While useful for
    determining the average properties of the stream (e.g., its mean distance
    track), a limited population of RR Lyrae stars are not sufficient to capture
    the fine-grained properties of streams expected to inform constraints on
    (e.g.) a population of perturbers in the galaxy. \JN{are there other
    methods?}
    % 
    Other methods for extracting stellar streams from the data include
    matched-filter approaches, where over-densities in color-magnitude space are
    inspected as possible clusters undergoing tidal stripping \JN{cite the many
    key papers using this method}. While the matched-filter technique has been
    extremely successful at stream discovery, it is not equipped to characterize
    streams at the level of detailed necessary to inform precise constraints on
    the potential of the galaxy and its substructure.
    % 
    In order to address the need for a flexible model of stellar streams both in
    photometric and kinematic spaces, in this work we develop a new method for
    characterizing streams and quantifying the membership probability of
    possible stream stars. Our approach provides several key advantage over
    previous stream modeling efforts, specifically in its ability to jointly
    model the kinematic and photometric properties of streams simultaneously.
    Furthermore, our method is well-suited to fitting streams with only partial
    or incomplete phase-space or photometric observations, enabling a full
    characterization of streams insofar as the available data allows.

@@

        % Observational errors complicate this picture. Instead of a certain
        % membership calculation, we must build a probabilistic model, assigning
        % a likelihood that an observed star falls along the stream's orbit.

@@

        The paper is organized as follows.  In \autoref{sec:method} we introduce
        our method for characterizing stream tracks and computing stellar
        membership.  We discuss the data in \autoref{sub:method:framing_the_data} and
        build a probabilistic framework to model the stream and background in
        \autoref{sub:method:likelihood_setup}-\autoref{sub:methods:priors}.  In
        \autoref{sec:results_mock} we apply our method to simulated data to
        demonstrate the model with a known ground truth dataset, then to the
        stellar streams \stream{GD-1} (\autoref{sec:results_gd1}) and
        \stream{Pal\,5} (\autoref{sec:results_pal5}).  We compare our approach
        to characterizing stellar streams to other methods in
        \autoref{sec:comparison}.  We conclude in \autoref{sec:conclusions},
        discussing the results of this work and future directions.

@@

    We now introduce our method for characterizing stellar stream tracks and
    computing stream membership likelihoods from N-D phase-space measurements.
    % TODO: rephrase following sentence
    We develop a flexible Bayesian model framework, that allows arbitrary
    numbers of phase-space dimensions to be incorporated in the model and
    permits missing phase-space observations.  Below, we offer a concise
    overview of our approach, followed by a more in-depth technical explanation
    in subsequent sections.  We start, in \autoref{sub:method:framing_the_data}, by
    transforming the astrometric data into a stream-oriented on-sky reference
    frame ($\phi_1, \phi_2, \parallax, \mu_{\phi_1}^*, \mu_{\phi_2}, v_r)$, for
    each stream. In this frame, the stream track lies along $\phi_1$, and the
    stream density may be described by distributions in the other parameters as
    injective functions of $\phi_1$.
    % TODO: ⬆ I think a few more sentences explaining why this is important,
    % feature variations, etc?  However, previously we discussed how stream will
    % vary along the track, with changes in the width and possible gaps.  Also
    % observational effects, e.g the distance gradient changes the on-sky width
    % (should this be mentioned earlier?)
    Likewise, the background of the Galactic field may be described with
    $\phi_1$-conditioned distributions.
    % TODO: ⬆ I think this should be expanded.
    Non-astrometric coordinates, e.g. photometry and metallicity, are also
    modelled along the stream and background as functions of $\phi_1$.

@@

    \subsection{Framing the Data} \label{sub:method:framing_the_data}

@@

    \subsection{Likelihood Setup}\label{sub:method:likelihood_setup}

@@

        The mixture coefficients $\qth{f}$ are normalized s.t.
        \begin{equation}
            \sum_{\mrm{q}} \qth{f}(\phi_1) = 1 \qquad \forall \phi_1
        \end{equation}
        In practice this is enforced by defining
        \begin{equation} \label{eq:mixture_weight_normalization}
            f_{b}(\phi_1) = 1 - \sum_{\rm{q}\in \mbs{I_Q}\backslash \{b\}} \qth{f}(\phi_1),
        \end{equation}
        the summation over all models except for some ``background" index $b \in
        \mbs{I_Q}$.  We discuss the mixture weights further in
        \autoref{sub:method:mixture_weight}.

@@

    \subsection{Missing Phase-Space Observations} \label{sub:method:missing_data}

@@

    \vspace{10pt}
    \subsection{Determination of the Mixture Weight} \label{sub:method:mixture_weight}

        The mixture weight $f_q(\phi_1)$ as used in \autoref{eq:general_model}
        is characterized as the fraction of stars belonging to the mixture
        component $q$ over an increment $[\phi_1, \phi_1 + \dif{\phi_1}]$. Thus,
        the mixture weight encodes information about the linear density of the
        stream, which has been shown to depend on substructure in the galaxy
        (CITE-x). The dependence of the mixture weight on the linear density is
        captured by the expression
        \begin{equation}\label{eq:fraction_param}
            f_q\left(w_{i,0}\right) = \frac{N_q\left(\phi_1\right)}{N_{\rm tot}\left(\phi_1\right)},
        \end{equation}
        where $N_q\left(\phi_1\right)$ is the number of stars in component $q$
        within an interval $[\phi_1, \phi_1 + \dif{\phi_1}]$, and $N_{\rm tot}$
        is the total number of stars in the same band (including those in
        component $q$). The linear  density for component $q$ is then entirely
        specified by the function $N_q(\phi_1)$.

@@

    \subsection{Astrometric Model} \label{sub:method:astrometric_model}

        \subsubsection{On-Stream} \label{ssub:method:astrometric_model:on_stream}

@@

            As a MDN, we implement the Gaussian distribution parameters with a
            multi-layered feed-forward neural network. The number of layers
            required depends on the maximum dimensionality of the Gaussian, and
            whether the neural network employs regularization techniques like
            dropout. We use alternating layers of linear and tanh units. Details
            of the network architecture are discussed in
            \autoref{sub:methods:model_implementation}.

@@

        \subsubsection{Off-stream} \label{ssub:method:astrometric_model:off_stream}

            We now outline the off-stream astrometric model. We take a similar
            approach to the on-stream astrometric model discussed in
            \autoref{ssub:method:astrometric_model:on_stream}, though we
            typically do not characterize the backgrounds with a Gaussian
            distribution. Instead, we utilize a range of distributions (i.e.,
            exponential, skew-normal) discussed in \autoref{app:distributions}.
            The parameters of the user-specified background distributions are
            themselves neural network outputs, such that any $\phi_1$ ``slice"
            of the background should be characterized by some analytic density,
            but the evolution of the backgrounds with $\phi_1$ can be complex.
            Details of the network architecture are discussed in
            \autoref{sub:methods:model_implementation}.

@@

    \subsection{Photometric Model} \label{sub:method:photometric_model}

        \subsubsection{On-Stream} \label{sub:method:photometric_model:on_stream}

            In photometric coordinates we model the stream as a
            single-population isochrone with the possibility of a non-zero
            distance gradient. We accomplish this by modeling the distance of
            the isochrone as a neural network output, parametrized as a function
            of $\phi_1$. The result is a distance track along the stream,
            estimated from the variation in its isochrone with $\phi_1$.

@@

            The second term in \autoref{eq:isochrone_prior_components} --
            $\prior_{obs}(\nth{\mbs{m}} | \phi_1)$ -- encodes all observational
            effects in the photometric dataset. Some effects are easily modeled,
            e.g. instruments have well-characterized magnitude limits, both
            faint and bright. Other effects are not easily modeled, e.g, complex
            location-dependent completeness issues \citep{GaiaCompleteness}.
            Accounting for the completeness is very challenging. In practice we
            avoid the contribution of completeness to $\prior_{obs}$, masking
            data for which it is relevant. \Gaia, for instance, is complete from
            $12 < G < 17$ and has high completeness up to $G ~20$, thus we mask
            $G > 20$.  The masks are easily modeled as magnitude limits and
            replace the complex completeness model in $\prior_{obs}$.
            \autoref{sub:method:missing_data} explains how masking is
            implemented in the models, though in this case every feature is
            masked, not only a few feature dimensions. With every featured
            masked, the photometric model in
            \autoref{eq:photometric_probability} reduces from an
            $F_m$-dimensional Gaussian to a 0-D Gaussian, which is a null
            distribution and does not enter into the likelihood.  In photometric
            coordinates completeness depends on a stars apparent magnitude; vied
            in in astrometric coordinates, the completeness cannot be described
            so simply. \Gaia, for instance has scanning law residuals in the
            density field. When photometric coordinates are masked due to
            completeness, we do not mask the corresponding  astrometric
            features, so that the stars  membership probability is determined
            solely by the astrometric model. This practice allows us to include
            stars with good astrometric measurements, not penalizing those stars
            for which the photometric model is insufficient.  We examine the
            results to confirm that the scan pattern does not impact the
            membership distribution.

@@

            % TODO? expand this section.
            The last approach to the issue of the PDMF, and the one we take in
            this work, is to render it unimportant.  With fixed age,
            metallicity, and other isochrone model parameters, the primary
            purpose of the isochrone is to allow the model to determine the
            distance modulus of the stream. For this, only a portion of the
            stream need be modeled.  For streams with observable main sequence
            turn-offs, like \stream{GD-1} in \gaia, this portion of the
            isochrone has both many stars and a relatively small range in
            masses. Thus, we restrict the mass range of the isochrone such that
            we are modeling only a region of interest over which the PDMF does
            not contribute to the final result.
            % TODO? work in something to the effect of: The PDMF doesn't have an
            % enormous impact, regardless as the isochrone model is very thin
            % and stars close to the isochrone photometrically are generally
            % quite far astrometrically. This is an advantange of simultaneous
            % modeling, where errors in one model are balanced by other parts of
            % the model.

@@

        \subsubsection{Linking to Astrometrics: the Distance Modulus and the Parallax} \label{ssub:method:linking_to_astrometrics}

@@

        \subsubsection{Off-stream}
        \label{sub:method:photometric_model:off_stream}

            In this section we introduce our model for the distribution of
            backgrounds in photometric coordinates.  The distribution of stars
            in color (i.e., $g-r$) and magnitude (i.e., $g$) is a complicated
            function of location in the galaxy \TODO{(e.g. [cite])}. Even within
            a field centered on the stream of interest, the color-magnitude
            diagram (CMD) is a combination of stellar populations distributed
            over a range in distances. Because of this, we do not expect an
            analytic density to provide a realistic description of the data. In
            this work we take a data-driven approach, and represent the CMD
            using a normalizing flow model.

@@

            % The specific architecture of the photometric normalizing flow are
            % discussed in \autoref{sub:methods:model_implementation}. But
            It is easy to evaluate the log-likelihood a normalizing flow at a
            point $\mbs{m}$ when all feature dimensions are present. When one
            (or more) feature dimension is missing, however, evaluating the
            marginal log-likelihood involves integrals over the flow. While the
            procedure is straightforward it is computationally expensive. As an
            alternative, we conservatively set the log-likelihood to $0$, the
            maximum value, overemphasizing the flow's contribution to the
            background model. Conceptually, up-weighting the background model
            decreases the false-positive rate in stream membership
            identification, at the cost of increasing the false-negative rate.
            However, the intrinsic width of stellar streams tends to be quite
            small, with correspondingly small $\sigma_\mu$ (see
            \autoref{eq:photometric_probability}); thus the effective region of
            the isochrone model is well determined and we do not find the
            marginal log-likelihood approach to be impactful. Another approach,
            which we leave to a future work, is to construct normalizing flows
            of each marginal distribution, e.g. $q_{\rm flow}\left(\{m_i\}_{i
            \in \mbs{I}_F, i \neq {j, k}} | m_j, m_k, \phi_1\right)$, and select
            for each $\nth{\mbs{m}}$ the correct flow given the present
            features.

@@

    \subsection{Priors} \label{sub:methods:priors}

        In \autoref{sub:method:likelihood_setup} we introduced the likelihood
        setup. In \autoref{sub:method:photometric_model:on_stream} we added
        priors in the form of the isochrone's PDMF and photometric observational
        constraints. In this section we introduce and develop more priors that
        are important for training the BMDN's parameters' neural network(s).

        \subsubsection{Parameter Bounds} \label{ssub:methods:priors:parameter_bounds}

@@

        \subsubsection{Region Prior} \label{sub:methods:priors:track_region_prior}

            In \autoref{ssub:method:astrometric_model:on_stream} the stream
            model's parameters were unconstrained. Bounds on the parameters and
            the neural network architecture were introduced in
            \autoref{ssub:methods:priors:parameter_bounds} to ensure the initial
            state of the model and network were compatible with the optimizer.
            Now we provide known information about physical systems, particular
            the track of the stream.  Our density modeling is aimed at
            characterization of \textit{known} streams, not stream discovery.
            From previous studies, the broad track of a stream is known for
            significant portions of its length \citep[e.g. see the atlas
            in][]{Mateu2022}. It makes sense, then, to provide this information
            to the model. Care must be taken that this prior information does
            not dominate the model.
            % A good balance between these two goals is to have a prior that
            % only influences the track on large scales.
            We introduce a region prior to encourage the stream model to pass
            through a user-defined control region, while also remaining
            compatible with the data.  We refer to the prior as a
            ``split"-Normal prior: a Gaussian split and separated at the peak
            and pieced back together with a flat region connecting the halves.
            The un-normalized PDF is given by:
            \begin{small}
            \begin{equation}
                \pdf(\theta,\mu,\sigma,w) \propto \begin{cases} 
                   \Exp{-\frac{1}{2}\left(\frac{x-(\mu-w)}{\sigma}\right)^2} & \phantom{\mu - w <}\ x < \mu - w \\
                    1 & \mu - w \leq x < \mu + w \\
                    \Exp{-\frac{1}{2}\left(\frac{x-(\mu+w)}{\sigma}\right)^2} & \mu + w < x
                \end{cases}.
            \end{equation}\end{small}
            The piecewise PDF is smooth up to the first derivative and can be
            used with gradient-based optimizers.

@@

    \vspace{5pt}
    \subsection{Model Implementation, aka the code} \label{sub:methods:model_implementation}

@@

        The models permit user-defined priors to be passed at model
        instantiation.  Priors specific to a single model are contained on that
        model, while priors connecting two (or more) models are held on the
        mixture model object. As an example of the latter, see
        \autoref{ssub:method:linking_to_astrometrics}, which connects the
        distance component of the astrometric and photometric stream models.
